Quotient maps and configuration spaces of hard disks

### 2022-02-09 manuscript/main.tex
base config space and quotient space definitions added in sec.4

One natural notion of distance assigns to two configurations $\config{p}, \config{q} \in \Lambda(n)$ a distance equal to the sum of the disk displacements required to transform one into the other, or
\begin{equation}
d_\Lambda(\config{p},\config{q}) = \sum_{i=1}^{n} {\Vert \vec{p}_i - \vec{q}_i \Vert}
\label{eq:labeled_distance}
\end{equation}
where $\Vert \vec{p}_i - \vec{q}_i \Vert$ is the geodesic distance between the two positions of the $i$th disk. Figure \ref{fig:figure4} shows these displacements for two configurations sampled uniformly at random on the base configuration spaces for the square and hexagonal toruses. Here, $d_\Lambda$ is the sum of the lengths of the vectors pointing from one disk to the other. Observe that $d_\Lambda$ is sensitive to symmetry operations in the sense that applying translations, permutations or lattice symmetries to one of the configurations changes $d_\Lambda$. That said, $d_\Lambda$ satisfies the requirements of a metric on the base configuration space (identity of indiscernibles, symmetry, and the triangle inequality) with proofs provided in App.\ \ref{sec:metric_proofs}.

@@

Let $\mathcal{T}$, $\mathcal{P}$, $\mathcal{I}$ and $\mathcal{L}$ respectively be the sets of rigid translations, permutations of the disk labels, inversion about the origin, and symmetries of the tiling of the plane. Formally, a configuration $\config{q}$ is a translation of $\config{p}$ by $\vec{t}$ if $\vec{q}_i = \vec{p}_i + \vec{t}$ for all disk indices $i$. Given a permutation $\pi \in \mathcal{P}$, $\config{q}$ is a permutation of $\config{p}$ if $\vec{q}_i = \pi(\vec{p}_i)$ for all $i$. A configuration $\config{q}$ is the inversion of $\config{p}$ if $\vec{q}_i = -\vec{p}_i$ for all $i$. Finally, for any symmetry element $L \in \mathcal{L}$ with representation $\mat{L}$, a configuration $\config{q}$ is a symmetric copy of $\config{p}$ if $\vec{q}_i = \mat{L} \vec{p}_i$ for all $i$. Observe that the operations belonging to all of these groups are isometric as required to use Eq.\ \ref{eq:unlabeled_distance}. Table \ref{table:symmetries} shows the different symmetry groups by which the configuration space is quotiented in this work, and the corresponding distances between the configurations in Fig.\ \ref{fig:figure4}. The $d_{\Lambda / S}$ are computed by fixing the first configuration and generating all copies of the second configuration that only differ by the action of $\mathcal{S / T}$, i.e., the discrete symmetry elements. Finding the rigid translation $T \in \mathcal{T}$ that minimizes $d_\Lambda\{\config{p}, T[S(\config{q})]\}$ for $S \in \mathcal{S / T}$ is a global optimization problem that is handled by the Tabu search algorithm \cite{chelouah2000tabu,glover1989tabu}. Finally, $d_{\Lambda / \mathcal{S}}$ is reported as the minimum of these distances for all $S \in \mathcal{S / T}$.

>>>>

One natural notion of distance assigns to two configurations \red{on the base configuration space} $\config{p}, \config{q} \in \Lambda(n)$ a distance equal to the sum of the disk displacements required to transform one into the other, or
\begin{equation}
d_\Lambda(\config{p},\config{q}) = \sum_{i=1}^{n} {\Vert \vec{p}_i - \vec{q}_i \Vert}
\label{eq:labeled_distance}
\end{equation}
where $\Vert \vec{p}_i - \vec{q}_i \Vert$ is the geodesic distance between the two positions of the $i$th disk. Figure \ref{fig:figure4} shows these displacements for two configurations sampled uniformly at random on the base configuration spaces for the square and hexagonal toruses. Here, $d_\Lambda$ is the sum of the lengths of the vectors pointing from one disk to the other. Observe that $d_\Lambda$ is sensitive to symmetry operations in the sense that applying translations, permutations or lattice symmetries to one of the configurations changes $d_\Lambda$. That said, $d_\Lambda$ satisfies the requirements of a metric on the base configuration space (identity of indiscernibles, symmetry, and the triangle inequality) with proofs provided in App.\ \ref{sec:metric_proofs}.

@@

Let $\mathcal{T}$, $\mathcal{P}$, $\mathcal{I}$ and $\mathcal{L}$ respectively be the sets of rigid translations, permutations of the disk labels, inversion about the origin, and symmetries of the tiling of the plane. Formally, a configuration $\config{q}$ is a translation of $\config{p}$ by $\vec{t}$ if $\vec{q}_i = \vec{p}_i + \vec{t}$ for all disk indices $i$. Given a permutation $\pi \in \mathcal{P}$, $\config{q}$ is a permutation of $\config{p}$ if $\vec{q}_i = \pi(\vec{p}_i)$ for all $i$. A configuration $\config{q}$ is the inversion of $\config{p}$ if $\vec{q}_i = -\vec{p}_i$ for all $i$. Finally, for any symmetry element $L \in \mathcal{L}$ with representation $\mat{L}$, a configuration $\config{q}$ is a symmetric copy of $\config{p}$ if $\vec{q}_i = \mat{L} \vec{p}_i$ for all $i$. Observe that the operations belonging to all of these groups are isometric as required to use Eq.\ \ref{eq:unlabeled_distance}. \red{The three quotient spaces considered in this work are namely the base configuration space modulo translation $\Lambda/\mathcal{T}$, the base configuration space modulo translations, permutations of disk labels, and inversion $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I}\}$ and the base configuration space modulo translations, permutations of disk labels, inversion and lattice symmetries $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I} \bigcup \mathcal{L}\}$. For the purposes of simplicity in the notation, Tab.\ \ref{table:symmetries} groups the symmetry groups by which the base configuration space is quotiented into a single term $\mathcal{S}_{i}$. Table \ref{table:symmetries} also shows the distances between the configurations shown in Fig.\ \ref{fig:figure4} in the base configuration space and in the three quotient space considered.} The $d_{\Lambda / S}$ are computed by fixing the first configuration and generating all copies of the second configuration that only differ by the action of $\mathcal{S / T}$, i.e., the discrete symmetry elements. Finding the rigid translation $T \in \mathcal{T}$ that minimizes $d_\Lambda\{\config{p}, T[S(\config{q})]\}$ for $S \in \mathcal{S / T}$ is a global optimization problem that is handled by the Tabu search algorithm \cite{chelouah2000tabu,glover1989tabu}. Finally, $d_{\Lambda / \mathcal{S}}$ is reported as the minimum of these distances for all $S \in \mathcal{S / T}$.
